
---
title: "Reds_Comp_2025"
author: "Villanova Sports Analytics Club"
date: "16-02-2025"
output: html_document
---

This is the Villanova Sports Analytics Club submission ot the 2025 Sports Analytics Hackathon



Relevant packages
```{r}
library(tidyverse)
library(data.table)
library(lubridate)
library(Metrics)
library(xgboost)

```



Loading datasets
```{r}
# Load datasets
lahman <- read_csv("lahman_people.csv")
savant <- read_csv("savant_data_2021_2023.csv")
sample <- read_csv("sample_submission.csv")

sample$PLAYING_TIME <- NA

# Create 2024 dataset from sample submission file
df_2024 <- sample %>%
  mutate(batter = PLAYER_ID, pitcher = PLAYER_ID) %>%
  mutate(game_year = 2024)

# Append 2024 dataset to savant data
savant <- bind_rows(savant, df_2024)
```


## Combining Lahman data

```{r}
# Process pitchers' data from Lahman dataset
pitchers_lahman <- savant %>% 
    select(game_date, game_year, pitcher) %>% 
    group_by(game_year) %>% 
    mutate(szn_start = as.IDate(min(game_date))) %>%  # Ensure it's IDate
    mutate(szn_start = if_else(game_year == 2024, as.IDate("2024-03-28"), szn_start)) %>% 
    group_by(szn_start, pitcher) %>%
    dplyr::slice(1) %>%
    ungroup() %>% 
    left_join(lahman, by = c("pitcher" = "player_mlb_id")) %>% 
    select(game_year, pitcher, szn_start, debut, birthDate)

# Process batters' data from Lahman dataset
batters_lahman <- savant %>% 
    select(game_date, game_year, batter) %>% 
    group_by(game_year) %>% 
    mutate(szn_start = as.IDate(min(game_date))) %>% 
    mutate(szn_start = if_else(game_year == 2024, as.IDate("2024-03-28"), szn_start)) %>%
    group_by(szn_start, batter) %>%
    dplyr::slice(1) %>%
    ungroup() %>% 
    left_join(lahman, by = c("batter" = "player_mlb_id")) %>% 
    select(game_year, batter, szn_start, debut, birthDate)

```

## drop missing & calc age, years in league, rookie
```{r}

pitchers_lahman <- pitchers_lahman %>% 
    filter(!is.na(debut)) %>% 
    mutate(age = floor(interval(birthDate, szn_start) / years(1)),
           years_in_league = as.numeric(game_year) - year(as.Date(debut)))

batters_lahman <- batters_lahman %>% 
    filter(!is.na(debut)) %>% 
    mutate(age = floor(interval(birthDate, szn_start) / years(1)),
           years_in_league = as.numeric(game_year) - year(as.Date(debut)))
```

## final lahman dfs
```{r}
pitchers_lahman <- pitchers_lahman %>% 
    rename(id = pitcher) %>% select(-szn_start,-debut,-birthDate,-years_in_league)
batters_lahman <- batters_lahman %>% 
    rename(id = batter) %>% select(-szn_start,-debut,-birthDate,-years_in_league)

```





## Identifying PAs in Savant Data

```{r}
pa_flag_vec <- c(
  field_out = 1, nan = 0, strikeout = 1, double = 1, strikeout_double_play = 1,
  single = 1, force_out = 1, hit_by_pitch = 1, grounded_into_double_play = 1,
  home_run = 1, walk = 1, caught_stealing_2b = 0, sac_bunt = 1, triple = 1,
  sac_fly = 1, field_error = 1, double_play = 1, catcher_interf = 1, fielders_choice_out = 1,
  fielders_choice = 1, pickoff_1b = 0, other_out = 0, caught_stealing_home = 0, pickoff_caught_stealing_2b = 0,
  caught_stealing_3b = 0, sac_fly_double_play = 1, pickoff_caught_stealing_home = 0, pickoff_2b = 0, run = 0,
  triple_play = 1, batter_interference = 1, pickoff_3b = 0, sac_bunt_double_play = 1, pickoff_caught_stealing_3b = 0
)


savant$pa_flag <- ifelse(
  savant$events %in% names(pa_flag_vec) & pa_flag_vec[savant$events] == 1, 
  1, 
  0
)



```


## calculating woba values
```{r}
# Apply weights manually if needed
weights <- c(single = 0.882, double = 1.254, triple = 1.590, home_run = 2.050, walk = 0.689, hit_by_pitch = 0.720, sac_fly = 0.25)

savant <- savant %>%
  mutate(
    weighted_value = case_when(
      events == "single" ~ weights["single"],
      events == "double" ~ weights["double"],
      events == "triple" ~ weights["triple"],
      events == "home_run" ~ weights["home_run"],
      events == "walk" ~ weights["walk"],
      events == "hit_by_pitch" ~ weights["hit_by_pitch"],
      events == "sac_fly" ~ weights["sac_fly"],
      TRUE ~ 0
    )
  )


```




## lineup order feature
```{r}
lineup_order <- savant %>%
    group_by(game_pk) %>% 
    filter(at_bat_number %in% 1:18, 
           pitch_number == 1) %>% 
    ungroup() %>% 
    group_by(game_pk, inning_topbot, batter) %>%  
    summarise(first_ab = min(at_bat_number, na.rm = TRUE), .groups = "drop") %>% 
    arrange(game_pk, inning_topbot, first_ab) %>%  
    group_by(game_pk, inning_topbot) %>% 
    mutate(lineup_spot = row_number()) %>%  
    ungroup()

# Calculate most frequent lineup spot by game_year and batter
most_frequent_lineup_spot <- lineup_order %>%
    left_join(savant %>% select(game_pk, game_year), by = "game_pk") %>%  # Add game_year to lineup_order
    group_by(batter, game_year, lineup_spot) %>%  
    tally() %>% 
    arrange(batter, game_year, desc(n)) %>%  
    slice_max(n, n = 1, with_ties = FALSE) %>% 
    select(-n) %>% 
    rename(prev_most_frequent_lineup_spot = lineup_spot)


most_frequent_lineup_spot$game_year = most_frequent_lineup_spot$game_year+1



```


## Creating season level woba, PAs and splits
```{r}
batters_L <- savant %>%
  select(game_year, batter, pa_flag, weighted_value, woba_denom, p_throws) %>% 
  filter(p_throws == "L") %>%
  group_by(game_year, batter) %>%
  summarise(
    PAs_LHP = sum(pa_flag),
    wOBA_LHP = sum(weighted_value, na.rm = TRUE) / sum(woba_denom, na.rm = TRUE)
  ) %>%
  ungroup()

batters_R <- savant %>%
  select(game_year, batter, pa_flag, weighted_value, woba_denom, p_throws) %>% 
  filter(p_throws == "R") %>%
  group_by(game_year, batter) %>%
  summarise(
    PAs_RHP = sum(pa_flag),
    wOBA_RHP = sum(weighted_value, na.rm = TRUE) / sum(woba_denom, na.rm = TRUE)
  ) %>%
  ungroup()

batters <- savant %>% 
  select(game_year, batter, pa_flag, stand, estimated_woba_using_speedangle, weighted_value, woba_denom) %>%
  group_by(game_year, batter) %>%
  summarise(
    PAs = sum(pa_flag),
    xWOBA = mean(estimated_woba_using_speedangle, na.rm = TRUE),
    wOBA = sum(weighted_value, na.rm = TRUE) / sum(woba_denom, na.rm = TRUE)
  ) %>%
  ungroup()

batters_with_R <- left_join(batters, batters_R, by = c("game_year", "batter"))
batters_full <- left_join(batters_with_R, batters_L, by = c("game_year", "batter"))


# Weighting previous seasons based on 5-3-2
# Define weight variables
w1 <- 5
w2 <- 3
w3 <- 2

batters_full_weighted <- batters_full %>%
  group_by(batter) %>%
  arrange(game_year) %>%
  mutate(
    weighted_PAs = {
      num <- w1 * coalesce(lag(PAs, 1), 0) +
             w2 * coalesce(lag(PAs, 2), 0) +
             w3 * coalesce(lag(PAs, 3), 0)
      denom <- if_else(is.na(lag(PAs, 1)), 0, w1) +
               if_else(is.na(lag(PAs, 2)), 0, w2) +
               if_else(is.na(lag(PAs, 3)), 0, w3)
      if_else(denom == 0, NA_real_, num / denom)
    },
    weighted_xWOBA = {
      num <- w1 * coalesce(lag(xWOBA, 1), 0) +
             w2 * coalesce(lag(xWOBA, 2), 0) +
             w3 * coalesce(lag(xWOBA, 3), 0)
      denom <- if_else(is.na(lag(xWOBA, 1)), 0, w1) +
               if_else(is.na(lag(xWOBA, 2)), 0, w2) +
               if_else(is.na(lag(xWOBA, 3)), 0, w3)
      if_else(denom == 0, NA_real_, num / denom)
    },
    weighted_wOBA = {
      num <- w1 * coalesce(lag(wOBA, 1), 0) +
             w2 * coalesce(lag(wOBA, 2), 0) +
             w3 * coalesce(lag(wOBA, 3), 0)
      denom <- if_else(is.na(lag(wOBA, 1)), 0, w1) +
               if_else(is.na(lag(wOBA, 2)), 0, w2) +
               if_else(is.na(lag(wOBA, 3)), 0, w3)
      if_else(denom == 0, NA_real_, num / denom)
    },
    weighted_PAs_LHP = {
      num <- w1 * coalesce(lag(PAs_LHP, 1), 0) +
             w2 * coalesce(lag(PAs_LHP, 2), 0) +
             w3 * coalesce(lag(PAs_LHP, 3), 0)
      denom <- if_else(is.na(lag(PAs_LHP, 1)), 0, w1) +
               if_else(is.na(lag(PAs_LHP, 2)), 0, w2) +
               if_else(is.na(lag(PAs_LHP, 3)), 0, w3)
      if_else(denom == 0, NA_real_, num / denom)
    },
    weighted_wOBA_LHP = {
      num <- w1 * coalesce(lag(wOBA_LHP, 1), 0) +
             w2 * coalesce(lag(wOBA_LHP, 2), 0) +
             w3 * coalesce(lag(wOBA_LHP, 3), 0)
      denom <- if_else(is.na(lag(wOBA_LHP, 1)), 0, w1) +
               if_else(is.na(lag(wOBA_LHP, 2)), 0, w2) +
               if_else(is.na(lag(wOBA_LHP, 3)), 0, w3)
      if_else(denom == 0, NA_real_, num / denom)
    },
    weighted_PAs_RHP = {
      num <- w1 * coalesce(lag(PAs_RHP, 1), 0) +
             w2 * coalesce(lag(PAs_RHP, 2), 0) +
             w3 * coalesce(lag(PAs_RHP, 3), 0)
      denom <- if_else(is.na(lag(PAs_RHP, 1)), 0, w1) +
               if_else(is.na(lag(PAs_RHP, 2)), 0, w2) +
               if_else(is.na(lag(PAs_RHP, 3)), 0, w3)
      if_else(denom == 0, NA_real_, num / denom)
    },
    weighted_wOBA_RHP = {
      num <- w1 * coalesce(lag(wOBA_RHP, 1), 0) +
             w2 * coalesce(lag(wOBA_RHP, 2), 0) +
             w3 * coalesce(lag(wOBA_RHP, 3), 0)
      denom <- if_else(is.na(lag(wOBA_RHP, 1)), 0, w1) +
               if_else(is.na(lag(wOBA_RHP, 2)), 0, w2) +
               if_else(is.na(lag(wOBA_RHP, 3)), 0, w3)
      if_else(denom == 0, NA_real_, num / denom)
    }
  ) %>%
  ungroup() %>%
  select(batter, game_year, PAs, starts_with("weighted"))

```


## Calculating defensive position counts
```{r}

# This gives the count of innings each player has started at each position in a year
savant_2 <- savant %>%
  select(game_date, game_year, inning, inning_topbot, game_pk, pitcher_1, 
         fielder_2, fielder_3, fielder_4, fielder_5, fielder_6, fielder_7, fielder_8, fielder_9) %>%
  filter(game_year<2024)

savant_2_filtered <- savant_2 %>%
  group_by(game_pk, inning, inning_topbot, game_year, .drop = FALSE) %>% 
  slice_head(n = 1) %>% # Selects the first row for each "Top" and "Bot"
  ungroup()


Position_counts <- savant_2_filtered %>%
  pivot_longer(
    cols = c(pitcher_1, starts_with("fielder_")), # Include pitcher_1 and all fielder columns
    names_to = "position",                       # New column indicating the position
    values_to = "player_id"                      # Column with the player ID
  ) %>%
  group_by(player_id, game_year, position) %>%
  summarize(count = n(), .groups = "drop") %>%  # Count occurrences
  pivot_wider(
    names_from = position, # Make positions columns (e.g., pitcher_1, fielder_2, etc.)
    values_from = count,   # Fill columns with counts
    values_fill = 0        # Replace NA with 0
  ) %>%
  arrange(player_id, game_year)

# Create empty 2024 rows
Position_2024 <- sample %>%
  rename(player_id = PLAYER_ID) %>%
  mutate(game_year = 2024) %>%
  left_join(Position_counts %>% select(player_id) %>% distinct(), by = "player_id") %>%
  mutate(across(starts_with("fielder_"), ~0), pitcher_1 = 0)

# Append 2024 Data Back
Position_counts <- bind_rows(Position_counts, Position_2024)


# Add lagged features for position counts
Position_counts_with_lags <- Position_counts %>%
  group_by(player_id) %>%
  arrange(game_year) %>%  # Ensure data is sorted by year for each player
  mutate(
    across(
      starts_with("fielder_"), 
      ~lag(.), 
      .names = "prev_year_{.col}"  # Create lagged versions of fielder counts
    ),
    prev_year_pitcher_1 = lag(pitcher_1)  # Add lag for pitcher counts
  ) %>%
  ungroup() %>%
  select(player_id, game_year, starts_with("prev_year"))


Position_counts_with_lags <- Position_counts_with_lags %>%
  rowwise() %>%  # Ensure calculations are done per row
  mutate(prev_year_pos_played = sum(c_across(starts_with("prev_year_fielder_")) > 0, na.rm = TRUE)) %>%  
  ungroup() %>%
  select(player_id,game_year,prev_year_pitcher_1,prev_year_pos_played)



```


## Joining batting+positions
```{r}
batters_order_weighted <- left_join(batters_full_weighted,most_frequent_lineup_spot, by=c("batter","game_year"))
# Combining batting stats with position counts
batters_pos_weighted <- right_join(Position_counts_with_lags,batters_order_weighted, by=c("player_id"="batter", "game_year"))


```




## Defining pitcher roles
```{r}

pitcher_roles <- savant %>%
  group_by(game_year, pitcher, game_pk) %>%
  summarise(
    batters_faced = sum(pa_flag),  # Total batters faced in each appearance
    is_sp = max(sp_indicator)  # If started, it should be 1 for that game
  ) %>%
  ungroup() %>%
  group_by(game_year, pitcher) %>%
  summarise(
    avg_bf_per_game = mean(batters_faced, na.rm = TRUE),
    total_games = n(),
    games_started = sum(is_sp),
    games_relieved = total_games - games_started
  ) %>%
  ungroup()

pitcher_roles <- pitcher_roles %>%
  mutate(
    role = case_when(
      games_started / total_games >= 0.7 | avg_bf_per_game > 15 ~ "SP",
      games_started / total_games < 0.3 & avg_bf_per_game < 8 ~ "RP")
  )

pitcher_roles <- pitcher_roles %>%
  mutate(
    is_starter = ifelse(role == "SP", 1, 0),
    is_reliever = ifelse(role == "RP", 1, 0)) %>%
  select(game_year,pitcher,is_reliever,is_starter,avg_bf_per_game )




```


## pitch count, velo and spin_rate
```{r}

#pitcher per batter
pitchers_pitchcount <- savant %>%
  select(game_year, pitcher, pa_flag, p_throws) %>% 
  group_by(game_year, pitcher) %>%                # Group by year and pitcher
  summarise(
    BFs = sum(pa_flag, na.rm = TRUE),         # Sum up batters faced
    pitches_per_BF = n() / BFs                  
  ) %>%
  ungroup()

#median of top 10 velo
velocity <- savant %>%
    filter(!is.na(release_speed)) %>%  # Ensure no missing values
    group_by(game_year, pitcher) %>%
    slice_max(release_speed, n = 10, with_ties = FALSE) %>%  # Select top 10 fastest pitches
    summarise(
        median_top10_velocity = median(release_speed, na.rm = TRUE),  # Median of top 10 fastest
    ) %>%
    ungroup()
# median of top 10 spin rate
spin_rate <-savant %>%
    filter(!is.na(release_spin_rate),
           pitch_type %in% c("SL", "CU", "KC", "SV", "ST", "SC")) %>% #only breaking balls 
    group_by(game_year, pitcher) %>%
    slice_max(release_spin_rate, n = 10, with_ties = FALSE) %>%  #Select top 10 spin rates
    summarise(
        median_top10_spin_rate = median(release_spin_rate, na.rm = TRUE)) %>%
    ungroup()

```


## woba, BFs splits for pitchers
```{r}
pitchers_L <- savant %>%
  select(game_year, pitcher, pa_flag, weighted_value, woba_denom, stand) %>% 
  filter(stand == "L") %>%
  group_by(game_year, pitcher) %>%
  summarise(
    BFs_LHP = sum(pa_flag),            # Sum up plate appearances
    wOBA_LHP = sum(weighted_value,na.rm=TRUE)/sum(woba_denom, na.rm=TRUE)) %>%
  ungroup()

pitchers_R <- savant %>%
  select(game_year, pitcher, pa_flag, weighted_value, woba_denom, stand) %>% 
  filter(stand == "R") %>%
  group_by(game_year, pitcher) %>%
  summarise(
    BFs_RHP = sum(pa_flag),            # Sum up plate appearances
    wOBA_RHP = sum(weighted_value,na.rm=TRUE)/sum(woba_denom, na.rm=TRUE)) %>%
  ungroup()

pitcher_woba <- savant %>% 
  select(game_year, pitcher, pa_flag, p_throws, estimated_woba_using_speedangle, weighted_value, woba_denom) %>% 
  group_by(game_year, pitcher) %>%
  summarise(
    BFs = sum(pa_flag),            # Sum up plate appearances
    xWOBA = mean(estimated_woba_using_speedangle,na.rm=TRUE),
    wOBA = sum(weighted_value,na.rm = TRUE)/sum(woba_denom,na.rm = TRUE)) %>%
  ungroup()



pitchers_with_R <- left_join(pitcher_woba,pitchers_R, by=c("game_year","pitcher"))

pitchers <- left_join(pitchers_with_R,pitchers_L, by=c("game_year","pitcher") )

# Define weight variables
w1 <- 5
w2 <- 3
w3 <- 2

pitchers_full_weighted <- pitchers %>%
  group_by(pitcher) %>%
  arrange(game_year) %>%
  mutate(
    weighted_BFs = {
      num <- w1 * coalesce(lag(BFs, 1), 0) +
             w2 * coalesce(lag(BFs, 2), 0) +
             w3 * coalesce(lag(BFs, 3), 0)
      denom <- if_else(is.na(lag(BFs, 1)), 0, w1) +
               if_else(is.na(lag(BFs, 2)), 0, w2) +
               if_else(is.na(lag(BFs, 3)), 0, w3)
      if_else(denom == 0, NA_real_, num / denom)
    },
    weighted_xWOBA = {
      num <- w1 * coalesce(lag(xWOBA, 1), 0) +
             w2 * coalesce(lag(xWOBA, 2), 0) +
             w3 * coalesce(lag(xWOBA, 3), 0)
      denom <- if_else(is.na(lag(xWOBA, 1)), 0, w1) +
               if_else(is.na(lag(xWOBA, 2)), 0, w2) +
               if_else(is.na(lag(xWOBA, 3)), 0, w3)
      if_else(denom == 0, NA_real_, num / denom)
    },
    weighted_wOBA = {
      num <- w1 * coalesce(lag(wOBA, 1), 0) +
             w2 * coalesce(lag(wOBA, 2), 0) +
             w3 * coalesce(lag(wOBA, 3), 0)
      denom <- if_else(is.na(lag(wOBA, 1)), 0, w1) +
               if_else(is.na(lag(wOBA, 2)), 0, w2) +
               if_else(is.na(lag(wOBA, 3)), 0, w3)
      if_else(denom == 0, NA_real_, num / denom)
    },
    weighted_BFs_LHP = {
      num <- w1 * coalesce(lag(BFs_LHP, 1), 0) +
             w2 * coalesce(lag(BFs_LHP, 2), 0) +
             w3 * coalesce(lag(BFs_LHP, 3), 0)
      denom <- if_else(is.na(lag(BFs_LHP, 1)), 0, w1) +
               if_else(is.na(lag(BFs_LHP, 2)), 0, w2) +
               if_else(is.na(lag(BFs_LHP, 3)), 0, w3)
      if_else(denom == 0, NA_real_, num / denom)
    },
    weighted_wOBA_LHP = {
      num <- w1 * coalesce(lag(wOBA_LHP, 1), 0) +
             w2 * coalesce(lag(wOBA_LHP, 2), 0) +
             w3 * coalesce(lag(wOBA_LHP, 3), 0)
      denom <- if_else(is.na(lag(wOBA_LHP, 1)), 0, w1) +
               if_else(is.na(lag(wOBA_LHP, 2)), 0, w2) +
               if_else(is.na(lag(wOBA_LHP, 3)), 0, w3)
      if_else(denom == 0, NA_real_, num / denom)
    },
    weighted_BFs_RHP = {
      num <- w1 * coalesce(lag(BFs_RHP, 1), 0) +
             w2 * coalesce(lag(BFs_RHP, 2), 0) +
             w3 * coalesce(lag(BFs_RHP, 3), 0)
      denom <- if_else(is.na(lag(BFs_RHP, 1)), 0, w1) +
               if_else(is.na(lag(BFs_RHP, 2)), 0, w2) +
               if_else(is.na(lag(BFs_RHP, 3)), 0, w3)
      if_else(denom == 0, NA_real_, num / denom)
    },
    weighted_wOBA_RHP = {
      num <- w1 * coalesce(lag(wOBA_RHP, 1), 0) +
             w2 * coalesce(lag(wOBA_RHP, 2), 0) +
             w3 * coalesce(lag(wOBA_RHP, 3), 0)
      denom <- if_else(is.na(lag(wOBA_RHP, 1)), 0, w1) +
               if_else(is.na(lag(wOBA_RHP, 2)), 0, w2) +
               if_else(is.na(lag(wOBA_RHP, 3)), 0, w3)
      if_else(denom == 0, NA_real_, num / denom)
    }
  ) %>%
  ungroup() %>%
  select(pitcher, game_year, BFs, starts_with("weighted"))



```


## Combing pitcher datasets
```{r}

# Combine all pitcher datasets
combined_pitcher_data <- pitchers_pitchcount %>%
  left_join(velocity, by = c("game_year", "pitcher")) %>%
  left_join(spin_rate, by = c("game_year", "pitcher")) %>%
  left_join(pitcher_roles, by = c("game_year", "pitcher"))

# Create Lagged Features
combined_pitcher_data_with_lags <- combined_pitcher_data %>%
  group_by(pitcher) %>%
  arrange(game_year) %>%  # Ensure data is sorted by year for each pitcher
  mutate(
    prev_year_pitches_per_BF = lag(pitches_per_BF),
    prev_year_median_top10_velocity = lag(median_top10_velocity),
    prev_year_median_top10_spin_rate = lag(median_top10_spin_rate),
    prev_year_is_starter = lag(is_starter),
    prev_year_is_reliever = lag(is_reliever),
    prev_year_avg_bf_per_game = lag(avg_bf_per_game)
  ) %>%
  ungroup() %>%
  select(
    game_year,
    pitcher,
    starts_with("prev_year_")  # Keep only lagged columns
  )


#join lagged features with BFs and woba
Pitchers_full <- left_join(pitchers_full_weighted,combined_pitcher_data_with_lags, by=c("game_year","pitcher"))


```





## Joining features with Lahman data and removing outliers
```{r}
# Combining batters final with lahman

batters_comb <- left_join(batters_pos_weighted, batters_lahman, by=c("player_id"="id", "game_year")) %>%
                drop_na(weighted_PAs, PAs)  # Drop only relevant columns
pitchers_comb <- left_join(Pitchers_full, pitchers_lahman, by=c("pitcher"="id", "game_year")) %>%
                drop_na(weighted_BFs, BFs)

batters_comb <- batters_comb %>%  mutate(across(everything(), ~ ifelse(is.infinite(.), NA, .)))
pitchers_comb <- pitchers_comb %>%  mutate(across(everything(), ~ ifelse(is.infinite(.), NA, .)))



batters_2024 <- batters_comb %>% filter(game_year==2024, coalesce(prev_year_pitcher_1, 0) < 5)


pitchers_2024 <- pitchers_comb %>% filter(game_year==2024, weighted_BFs>1)


# removing low PA counts
batters_final <- batters_comb %>% filter(weighted_PAs>200,
                                          prev_year_pitcher_1<5,
                                          game_year<2024)

pitchers_final <- pitchers_comb %>% filter(weighted_BFs>150,
                                            game_year<2024)

```

## Batters model
```{r}
set.seed(123)
# Train-Test Split
player_ids <- unique(batters_final$player_id)
train_player_ids <- sample(unique(batters_final$player_id), size = round(0.8 * length(unique(batters_final$player_id))), replace = FALSE)

train_df <- batters_final %>% filter(player_id %in% train_player_ids)
test_df  <- batters_final %>% filter(!player_id %in% train_player_ids)

# Preparing Data Matrices
train_matrix <- xgb.DMatrix(
  data = as.matrix(train_df %>% select(-c(player_id, PAs, game_year))),
  label = train_df$PAs
)
test_matrix <- xgb.DMatrix(
  data = as.matrix(test_df %>% select(-c(player_id, PAs, game_year))),
  label = test_df$PAs
)

param_grid <- expand.grid(
  max_depth = c(3, 4, 5),
  eta = c(0.01, 0.05),
  min_child_weight = c(5, 10),
  subsample = c(0.5, 0.7),
  colsample_bytree = c(0.5, 0.7),
  lambda = c(2, 5, 10),
  alpha = c(1, 5, 10)
)

# Hyperparameter Grid Search
# param_grid <- expand.grid(
#   max_depth = c(3, 5, 7),
#   eta = c(0.01, 0.05, 0.1),
#   min_child_weight = c(1, 3, 5),
#   subsample = c(0.7, 0.8, 1.0),
#   colsample_bytree = c(0.7, 0.8, 1.0),
#   lambda = c(0.5, 1, 1.5),
#   alpha = c(0, 0.5, 1)
# )

cv_results <- lapply(1:nrow(param_grid), function(i) {
  params <- as.list(param_grid[i, ])
  params$objective <- "reg:squarederror"
  params$eval_metric <- "rmse"
  
  cv <- xgb.cv(
    params = params,
    data = train_matrix,
    nrounds = 1000,
    nfold = 5,
    early_stopping_rounds = 50,
    verbose = FALSE
  )
  
  list(
    params = params,
    best_rmse = min(cv$evaluation_log$test_rmse_mean),
    best_iteration = which.min(cv$evaluation_log$test_rmse_mean)
  )
})

best_result <- cv_results[[which.min(sapply(cv_results, function(x) x$best_rmse))]]
cat("Best Parameters:\n")
print(best_result$params)
cat("Best RMSE:", best_result$best_rmse, "\n")
cat("Best Iteration:", best_result$best_iteration, "\n")

# Final Model Training
best_params <- best_result$params
best_params$objective <- "reg:squarederror"
best_params$eval_metric <- "rmse"

```

```{r}
final_model <- xgb.train(
  params = best_params,
  data = train_matrix,
  nrounds = best_result$best_iteration,
  watchlist = list(train = train_matrix, test = test_matrix),
  verbose = 1
)



# Predictions and Evaluation
final_preds <- predict(final_model, test_matrix)
final_rmse <- sqrt(mean((test_df$PAs - final_preds)^2))
cat("Final Test RMSE:", final_rmse, "\n")

# Feature Importance
importance_matrix <- xgb.importance(model = final_model)
xgb.plot.importance(importance_matrix, top_n = 25, measure = "Gain")




# Generate importance matrix
importance_matrix <- xgb.importance(model = final_model)

# Convert to data frame for ggplot
importance_df <- importance_matrix  # Select top 25 features

# Create a custom ggplot importance plot
ggplot(importance_df, aes(x = reorder(Feature, Gain), y = Gain)) +
  geom_bar(stat = "identity", fill = "red4", alpha = 0.8) +  # Professional color scheme
  coord_flip() +  # Horizontal bars for readability
  ggtitle("Feature Importance (Batters)") +
  theme_minimal(base_size = 14) +  # Clean theme
  theme(
    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),  # Centered, bold title
    axis.title.x = element_text(size = 14, face = "bold"),  # Larger x-axis label
    axis.title.y = element_text(size = 14, face = "bold"),  # Larger y-axis label
    axis.text = element_text(size = 12),  # Increase tick label size
    panel.grid.major = element_line(color = "grey80"),  # Subtle grid lines
    panel.grid.minor = element_blank()
  ) +
  labs(x = "Features", y = "Feature Importance (Gain)")



ggplot(test_df, aes(x = PAs, y = final_preds)) +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Actual vs. Predicted BFs", x = "Actual BFs", y = "Predicted BFs") +
  theme_minimal()

```



## Secondary low sample batters model
```{r}
# Remove NAs before training the model
batters_comb_clean <- batters_comb %>% filter(prev_year_pitcher_1<5,
                                          game_year<2024) %>% drop_na(age, weighted_PAs, PAs) 

# Train secondary model
secondary_model_batters <- lm(PAs ~ age + 
                        weighted_PAs, data = batters_comb_clean)


summary(secondary_model_batters)


# Filter small sample players
small_sample_df <- batters_comb %>% 
  filter(weighted_PAs < 200, game_year < 2024, prev_year_pitcher_1 < 5) %>%
  drop_na(age, weighted_PAs)  # Drop NAs before predicting

# Make predictions
small_sample_df <- small_sample_df %>%
  mutate(predicted_PAs = predict(secondary_model_batters, newdata = small_sample_df))

# Compute RMSE
rmse <- sqrt(mean((small_sample_df$PAs - small_sample_df$predicted_PAs)^2, na.rm = TRUE))


# Print RMSE
cat("Final Test RMSE:", rmse, "\n") 

```





## Predicting 2024 Batters
```{r}
# Ensure df_2024 has only necessary features
new_data_matrix <- xgb.DMatrix(
  data = as.matrix(batters_2024 %>% select(-c(player_id, game_year,PAs)))
)

batters_2024 <- batters_2024 %>% mutate(predicted_PAs = predict(final_model, new_data_matrix),
                                        low_sample_predicted_PAs = 
                                          predict(secondary_model_batters, batters_2024))

```


# Pitchers model
```{r}
set.seed(123)
# Train-Test Split
player_ids <- unique(pitchers_final$pitcher)
train_player_ids <- sample(unique(pitchers_final$pitcher), size = round(0.8 * length(unique(pitchers_final$pitcher))), replace = FALSE)

train_df <- pitchers_final %>% 
  filter(pitcher %in% train_player_ids)
test_df  <- pitchers_final %>% filter(!pitcher %in% train_player_ids)



# Preparing Data Matrices with Weights
train_matrix <- xgb.DMatrix(
  data = as.matrix(train_df %>% select(-c(pitcher, BFs, game_year))), 
  label = train_df$BFs,
)

test_matrix <- xgb.DMatrix(
  data = as.matrix(test_df %>% select(-c(pitcher, BFs, game_year))),
  label = test_df$BFs
)


# Hyperparameter Grid Search
# param_grid <- expand.grid(
#   max_depth = c(3, 5, 7),
#   eta = c(0.01, 0.05, 0.1),
#   min_child_weight = c(1, 3, 5),
#   subsample = c(0.7, 0.8, 1.0),
#   colsample_bytree = c(0.7, 0.8, 1.0),
#   lambda = c(0.5, 1, 1.5),
#   alpha = c(0, 0.5, 1)
# )

# param_grid <- expand.grid(
#   max_depth = c(3, 5),
#   eta = c(0.05, 0.1),
#   min_child_weight = c(1, 5),
#   subsample = c(0.8, 1.0),
#   colsample_bytree = c(0.8, 1.0),
#   lambda = c(0.5, 1, 1.5),
#   alpha = c(0, 0.5, 1)
# )
param_grid <- expand.grid(
  max_depth = c(3, 4, 5),
  eta = c(0.01, 0.05),
  min_child_weight = c(5, 10),
  subsample = c(0.5, 0.7),
  colsample_bytree = c(0.5, 0.7),
  lambda = c(2, 5, 10),
  alpha = c(1, 5, 10)
)

cv_results <- lapply(1:nrow(param_grid), function(i) {
  params <- as.list(param_grid[i, ])
  params$objective <- "reg:squarederror"
  params$eval_metric <- "rmse"
  
  cv <- xgb.cv(
    params = params,
    data = train_matrix,
    nrounds = 1000,  # ⬆ More rounds to accommodate lower eta
    nfold = 5,
    early_stopping_rounds = 50,  # ⬆ Increase early stopping patience
    verbose = FALSE
  )
  
  list(
    params = params,
    best_rmse = min(cv$evaluation_log$test_rmse_mean),
    best_iteration = which.min(cv$evaluation_log$test_rmse_mean)
  )
})

best_result <- cv_results[[which.min(sapply(cv_results, function(x) x$best_rmse))]]

cat("Best Parameters:\n")
print(best_result$params)
cat("Best RMSE:", best_result$best_rmse, "\n")
cat("Best Iteration:", best_result$best_iteration, "\n")

# Final Model Training
best_params <- best_result$params
best_params$objective <- "reg:squarederror"
best_params$eval_metric <- "rmse"

final_model <- xgb.train(
  params = best_params,
  data = train_matrix,
  nrounds = best_result$best_iteration,
  watchlist = list(train = train_matrix, test = test_matrix),
  verbose = 1
)

```

```{r}
final_model <- xgb.train(
  params = best_params,
  data = train_matrix,
  nrounds = best_result$best_iteration,
  watchlist = list(train = train_matrix, test = test_matrix),
  verbose = 1
)

# Predictions and Evaluation
final_preds <- predict(final_model, test_matrix)
final_rmse <- sqrt(mean((test_df$BFs - final_preds)^2))
cat("Final Test RMSE:", final_rmse, "\n")

# Feature Importance
importance_matrix <- xgb.importance(model = final_model)
xgb.plot.importance(importance_matrix, top_n = 25, measure = "Gain")


# Generate importance matrix
importance_matrix <- xgb.importance(model = final_model)

# Convert to data frame for ggplot
importance_df <- importance_matrix  # Select top 25 features

# Create a custom ggplot importance plot
ggplot(importance_df, aes(x = reorder(Feature, Gain), y = Gain)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.8) +  # Professional color scheme
  coord_flip() +  # Horizontal bars for readability
  ggtitle("Feature Importance (Pitchers)") +
  theme_minimal(base_size = 14) +  # Clean theme
  theme(
    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),  # Centered, bold title
    axis.title.x = element_text(size = 14, face = "bold"),  # Larger x-axis label
    axis.title.y = element_text(size = 14, face = "bold"),  # Larger y-axis label
    axis.text = element_text(size = 12),  # Increase tick label size
    panel.grid.major = element_line(color = "grey80"),  # Subtle grid lines
    panel.grid.minor = element_blank()
  ) +
  labs(x = "Features", y = "Feature Importance (Gain)")


ggplot(test_df, aes(x = BFs, y = final_preds)) +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Actual vs. Predicted BFs", x = "Actual BFs", y = "Predicted BFs") +
  theme_minimal()


```




# Low Sample Model Pitchers
```{r}
# Remove NAs before training the model
pitchers_comb_clean <- pitchers_comb %>% filter(weighted_BFs>0, game_year<2024) %>%
  drop_na(age, BFs)

# Train secondary model
secondary_model_pitchers <- lm(BFs ~ age + weighted_BFs, 
                        data = pitchers_comb_clean)


summary(secondary_model_pitchers)


# Filter small sample players
small_sample_df <- pitchers_comb %>% 
  filter(weighted_BFs < 150, game_year < 2024) %>%
  drop_na(age, weighted_BFs)  # Drop NAs before predicting

# Make predictions
small_sample_df <- small_sample_df %>%
  mutate(predicted_BFs = predict(secondary_model_pitchers, newdata = small_sample_df))

# Compute RMSE
rmse <- sqrt(mean((small_sample_df$BFs - small_sample_df$predicted_BFs)^2, na.rm = TRUE))


# Print RMSE
cat("Final Test RMSE:", rmse, "\n")

```


## Predicting for 2024 Pitchers
```{r}
# Ensure df_2024 has only necessary features
new_data_matrix <- xgb.DMatrix(
  data = as.matrix(pitchers_2024 %>% select(-c(pitcher, game_year,BFs)))
)


pitchers_2024 <- pitchers_2024 %>% mutate(predicted_BFs = predict(final_model, new_data_matrix),
                                        low_sample_predicted_BFs = 
                                          predict(secondary_model_pitchers, pitchers_2024))



```





## Making final df for submission
```{r}
#
batters_2024$predicted_PAs <- ifelse(batters_2024$weighted_PAs<200, 
                                     batters_2024$low_sample_predicted_PAs, 
                                     batters_2024$predicted_PAs)

batters_2024$predicted_PAs <- ifelse(batters_2024$predicted_PAs<0, 
                                     batters_2024$weighted_PAs, 
                                     batters_2024$predicted_PAs)

batters_2024_final <- batters_2024 %>% select(player_id,predicted_PAs)

pitchers_2024$predicted_BFs <- ifelse(pitchers_2024$weighted_BFs<150, 
                                     pitchers_2024$low_sample_predicted_BFs, 
                                     pitchers_2024$predicted_BFs)

pitchers_2024$predicted_BFs <- ifelse(pitchers_2024$predicted_BFs<0, 
                                     pitchers_2024$weighted_BFs, 
                                     pitchers_2024$predicted_BFs)

pitchers_2024_final <- pitchers_2024 %>% select(pitcher,predicted_BFs)

submission_df <- sample %>% left_join(batters_2024_final,by=c("PLAYER_ID"="player_id"))
submission_df <- submission_df %>% left_join(pitchers_2024_final,by=c("PLAYER_ID"="pitcher"))


submission_df$PLAYING_TIME <- pmax(submission_df$predicted_PAs, submission_df$predicted_BFs, na.rm = TRUE)


submission_df <- submission_df %>% select(PLAYER_ID,PLAYING_TIME)

write_csv(submission_df,"reds_submission_final.csv")


```



# Exporting dataset for graphs
```{r}

batter_age <- batters_comb %>% select(player_id,game_year,PAs,age,weighted_PAs)

pitcher_age <- pitchers_comb %>% select(pitcher,game_year,BFs,age)


write_csv(batter_age,"batter_age.csv")
write_csv(pitcher_age,"pitcher_age.csv")



```

